---
title: "CytoGLMM Workflow"
author:
  name: Christof Seiler
  affiliation: Department of Statistics, Stanford University
output:
  BiocStyle::html_document2:
    toc_float: true
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Goal

This is a step-by-step guide for a complete analysis of mass cytometry data.

# Prerequisites

Install [R](https://www.r-project.org/) and [RStudio](https://www.rstudio.com/). Open this `Rmd` file in RStudio. Then run the following code to install all required packages.

```{r install_packages, warning=FALSE, message=FALSE, eval=FALSE}
pkgs_needed = c("devtools","tidyverse","magrittr","FlowRepositoryR",
                "flowCore","openCyto","ggcyto","scales","parallel",
                "RColorBrewer","ggcorrplot","SummarizedExperiment",
                "lme4","lmerTest")
letsinstall = setdiff(pkgs_needed, installed.packages())
if (length(letsinstall) > 0) {
  source("http://bioconductor.org/biocLite.R")
  biocLite(letsinstall)
}
# package is still private
devtools::install_github("ChristofSeiler/CytoGLMM", auth_token = "7e51f0606cfec76eb1328812ac263f335c5f285f")
```

Load packages.

```{r load_packages}
library("CytoGLMM")
library("tidyverse")
library("magrittr")
library("FlowRepositoryR")
library("flowCore")
library("openCyto")
library("ggcyto")
library("scales")
library("parallel")
library("RColorBrewer")
library("ggcorrplot")
library("SummarizedExperiment")
library("lme4")
library("lmerTest")
```

Set plotting style and assign computational resources.

```{r set_parameters}
theme_set(theme_light())
ncores = parallel::detectCores()
```

# Download and Prepare Data

## FCS Files

Reanalysis of mass cytometry data from @aghaeepour2017immune. We use the package `FlowRepositoryR` to download `fcs` files from [FlowRepository](http://flowrepository.org/id/FR-FCM-ZY3Q). We download mass cytometry data measured on whole blood samples collected from 16 women during pregnancy stimulated with IFNa at first and third trimester.

```{r download_data, eval=FALSE}
repo = flowRep.get("FR-FCM-ZY3Q")
data = download(repo, only.files = ".*_2_IFNa.*fcs")
data = download(repo, only.files = ".*_BL_IFNa.*fcs")
```

## Sample Table

The previous commands downloaded `fcs` files to the `FR-FCM-ZY3Q` folder. Now, we prepare the sample table by parsing the `fcs` filenames. For some studies, it might be easier to prepare a `sample_table.csv` textfile and load it using the `readr` package.

```{r read_csv}
fcs_files = list.files(path = "FR-FCM-ZY3Q/", pattern = "fcs")
map_time = function(x) {
  if (str_detect(x, "_2_")) "3rd trimester"
  else if (str_detect(x, "_BL_")) "1st trimester"
  else NA
}
sample_table = tibble(
  donor = str_extract(fcs_files, "PTLG[0-9]{3}"),
  term = sapply(fcs_files, map_time) %>% as.factor,
  file_name = paste0("FR-FCM-ZY3Q/",fcs_files)
)
sample_table
```

## Marker Table

Load marker isotopes and protein names. Make sure that marker names don't have `-` or start with a number. This is important to be compatible with the formula syntax of `R`.

```{r load_markers}
markers = read_csv("markers.csv")
markers$protein_name %<>% str_replace_all("-","_")
markers$protein_name %<>% make.names
markers
```

Check if bead normalized. This only works if the bead normalizer added an additional column `beadDist` to the `fcs`.

```{r bead}
ifelse(sum(colnames(read.FCS(sample_table$file_name[1])) == "beadDist")==1,
       yes = "bead normalized",
       no = "not bead normalized")
```

# Gating

Gate with [openCyto](https://bioconductor.org/packages/release/bioc/html/openCyto.html) according to supplementary material (Fig. S1) from @aghaeepour2017immune.

```{r opencyto, fig.wide=TRUE}
# load data
ncfs = read.ncdfFlowSet(sample_table$file_name, mc.cores = ncores)
gs = GatingSet(ncfs)
pData(gs) = cbind(pData(gs),sample_table)
trans_func = function(x) asinh(x/5)
inv_func = function(x) 5*sinh(x)
trans_obj = trans_new("asinh_cytof", trans_func, inv_func)
translist = transformerList(markers$isotope[-1], trans_obj)
gs = transform(gs, translist)
# apply gating template
gt_aghaeepour = gatingTemplate("gating_template.csv")
plot(gt_aghaeepour)
gating(gt_aghaeepour, gs, mc.cores = ncores, parallel_type = "multicore")
# rename some subsets
setNode(gs, "/singlet/leukocyte/mononuclear/CD3+CD19-/CD4+CD8a-/CD45RA+", 
        "CD4+Tnaive") %>% invisible
setNode(gs, "/singlet/leukocyte/mononuclear/CD3+CD19-/CD4+CD8a-/CD45RA-", 
        "CD4+Tmem") %>% invisible
setNode(gs, "/singlet/leukocyte/mononuclear/CD3+CD19-/CD4-CD8a+/CD45RA+", 
        "CD8+Tnaive") %>% invisible
setNode(gs, "/singlet/leukocyte/mononuclear/CD3+CD19-/CD4-CD8a+/CD45RA-", 
        "CD8+Tmem") %>% invisible
setNode(gs, "CD3-CD19+", "B") %>% invisible
setNode(gs, "CD3+CD19-", "T") %>% invisible
setNode(gs, "CD4+CD8a-", "CD4+T") %>% invisible
setNode(gs, "CD4-CD8a+", "CD8+T") %>% invisible
setNode(gs, "CD7+", "NK") %>% invisible
setNode(gs, "CD14+CD16-", "cMC") %>% invisible
setNode(gs, "CD14-CD16+", "ncMC") %>% invisible
setNode(gs, "CD14+CD16+", "intMC") %>% invisible
plot(gs)
# hide nodes
nodes_to_hide = c(
  "granulocyte", "CD19+", "CD3+", "CD3+CD19+",
  "CD8a+", "CD4+", "CD16+", "CD14+", "CD14-CD16-",
  "CD4+CD8a+", "CD4+Tnaive/CD25+", "CD4+Tmem/CD25+", 
  "CD4+Tnaive/FoxP3+", "CD4+Tmem/FoxP3+"
  )
for(this_node in nodes_to_hide) setNode(gs, this_node, FALSE)
plot(gs)
```

Visualize gates with [ggcyto](https://bioconductor.org/packages/release/bioc/html/ggcyto.html).

```{r ggcyto}
set.seed(0xdada)
ids = sample(length(gs), size = 6)
ggcyto(gs[ids], aes(x = CD66, y = CD45), subset = "leukocyte") + 
  ggcyto_par_set(limits = "instrument") + geom_hex(bins = 64) + 
  geom_gate(c("mononuclear","granulocyte"))
ggcyto(gs[ids], aes(x = CD3, y = CD19), subset = "mononuclear") + 
  ggcyto_par_set(limits = "instrument") + geom_hex(bins = 64) +
  geom_gate(c("B","T","CD3-CD19-"))
ggcyto(gs[ids], aes(x = CD14, y = CD7), subset = "CD3-CD19-") + 
  ggcyto_par_set(limits = "instrument") + geom_hex(bins = 64) +
  geom_gate("NK")
ggcyto(gs[ids], aes(x = CD14, y = CD16), subset = "CD7-") + 
  ggcyto_par_set(limits = "instrument") + geom_hex(bins = 64) +
  geom_gate(c("cMC","ncMC","intMC"))
ggcyto(gs[ids], aes(x = CD4, y = CD8a), subset = "T") + 
  ggcyto_par_set(limits = "instrument") + geom_hex(bins = 64) +
  geom_gate(c("CD4+T","CD8+T","CD4-CD8a-"))
ggcyto(gs[ids], aes(x = TCRgd, y = CD3), subset = "CD4-CD8a-") + 
  ggcyto_par_set(limits = "instrument") + geom_hex(bins = 64) +
  geom_gate("gdT")
ggcyto(gs[ids], aes(x = CD4, y = CD45RA), subset = "CD4+T") + 
  ggcyto_par_set(limits = "instrument") + geom_hex(bins = 64) +
  geom_gate(c("CD4+Tnaive","CD4+Tmem"))
ggcyto(gs[ids], aes(x = FoxP3, y = CD25), subset = "CD4+Tnaive") + 
  ggcyto_par_set(limits = "instrument") + geom_hex(bins = 64) +
  geom_gate("Tregsnaive")
ggcyto(gs[ids], aes(x = FoxP3, y = CD25), subset = "CD4+Tmem") + 
  ggcyto_par_set(limits = "instrument") + geom_hex(bins = 64) +
  geom_gate("Tregsmem")
ggcyto(gs[ids], aes(x = CD8a, y = CD45RA), subset = "CD8+T") + 
  ggcyto_par_set(limits = "instrument") + geom_hex(bins = 64) +
  geom_gate(c("CD8+Tnaive","CD8+Tmem"))
```

Combine cell types of interest into one data frame. Change marker name from isotope name to protein name and remove markers that are unmapped.

```{r convert}
nodes_exclude = c("root","singlet","leukocyte","mononuclear",
                  "CD3-CD19-","CD7-","CD4-CD8a-",
                  "CD4+T","CD8+T","T")
nodes_all = getNodes(gs, path = "auto") 
nodes_select = nodes_all[!nodes_all %in% nodes_exclude]
df_samples = lapply(nodes_select, function(celltype) {
  fset = getData(gs, celltype)
  lapply(seq(fset), function(sample_id) {
    marker_ids = which(fset@colnames %in% markers$isotope)
    exprs = as_tibble(exprs(fset[[sample_id]]))[,marker_ids]
    file_name = pData(fset[sample_id])$file_name
    exprs %>% add_column(file_name, celltype)
  }) %>% bind_rows
}) %>% bind_rows
df_samples %<>% inner_join(sample_table,by = "file_name")
oldnames = markers$isotope
newnames = markers$protein_name
df_samples %<>% rename_at(vars(oldnames), ~ newnames)
str(df_samples)
```

List cell counts per donor and trimester.

```{r list_cells_per_donor}
table(df_samples$donor,df_samples$term)
```

Plot abundance of each celltype per sample.

```{r cell_abundance}
df_abundance = df_samples %>% 
  group_by(file_name, donor, term, celltype) %>% 
  tally()
ggplot(df_abundance, aes(term, n, color = term)) + 
  geom_violin() + 
  geom_jitter(width = 0.2, alpha = 0.5) + 
  facet_wrap(~celltype, nrow = 2) +
  theme(axis.text.x = element_blank())
```

Focus on functional proteins.

```{r select_proteins}
protein_names = markers %>% 
  dplyr::filter(type == "function") %>% 
  .$protein_name
protein_names
```

Declare the columns in `df_samples` that are not protein markers. In our example, we have donor ID, time point when the sample was collected, `FCS` filename, and the cell type that we have defined through gating.

```{r}
sample_info_names = c(names(sample_table),"celltype")
sample_info_names
```

# Data Exploration 

## MDS

MDS on median marker expression of all cell types.

```{r plot_mds_celltype, fig.wide=TRUE}
CytoGLMM::plot_mds(df_samples,
                   protein_names = protein_names,
                   sample_info_names = sample_info_names,
                   color = "celltype")
```

Subset to NK cells to illustrate visualization for one cell type.

```{r subset_nk_cells}
df_samples_subset = df_samples %>% dplyr::filter(celltype == "NK")
```

MDS on median marker expression of NK cells.

```{r plot_mds_nk, fig.wide=TRUE}
CytoGLMM::plot_mds(df_samples_subset,
                   protein_names = protein_names,
                   sample_info_names = sample_info_names,
                   color = "term")
```

## Heatmap

Heatmap of median marker expression of all cell types.

```{r plot_heatmap_celltype}
CytoGLMM::plot_heatmap(df_samples,
                       protein_names = protein_names,
                       sample_info_names = sample_info_names,
                       arrange_by_1 = "term",
                       arrange_by_2 = "celltype")
```

Heatmap of median marker expression of NK cells.

```{r plot_heatmap_nk}
CytoGLMM::plot_heatmap(df_samples_subset,
                       protein_names = protein_names,
                       sample_info_names = sample_info_names,
                       arrange_by_1 = "term")
```

## PCA

PCA plot of all cell types.

```{r plot_prcomp_celltype, fig.wide=TRUE}
CytoGLMM::plot_prcomp(df_samples,
                      protein_names = protein_names,
                      color_var = "celltype",
                      repel = TRUE)
```

PCA plot of NK cells.

```{r plot_prcomp_nk, fig.wide = TRUE}
CytoGLMM::plot_prcomp(df_samples_subset,
                      protein_names = protein_names,
                      color_var = "term",
                      repel = TRUE)
```

## Density Plots

Density plots of one marker for all donors.

```{r density_plot, fig.wide=TRUE}
ggplot(df_samples_subset, aes_string(x = "pCREB", color = "term")) + 
  geom_density() + 
  facet_wrap(~donor)
```

## Two-Dimensional Histograms

Two-dimensional histograms for plotting two markers for all donors.

```{r hist2_donors}
colorscale = scale_fill_gradientn(
  colors = rev(brewer.pal(9, "YlGnBu")), 
  values = c(0, exp(seq(-5, 0, length.out = 100)))
  )
ggplot(df_samples_subset, aes_string(x = "pSTAT1", y = "pSTAT3")) + 
  geom_hex(bins = 64) +
  colorscale + 
  coord_fixed() +
  facet_wrap(~donor)
```

Two-dimensional histograms for group comparisons.

```{r hist2_groups}
ggplot(df_samples_subset, aes_string(x = "pSTAT1", y = "pSTAT3")) + 
  geom_hex(bins = 64) +
  colorscale + 
  coord_fixed() +
  facet_wrap(~term)
```

NK cell count. List the smallest and largest.

```{r cell_count}
df_samples_subset %>% group_by(term,donor) %>% tally %>% arrange(n)
df_samples_subset %>% group_by(term,donor) %>% tally %>% arrange(desc(n))
```

## Marker Correlations

Plot marker correlations.

```{r marker_correlations}
mcor = cor(df_samples_subset %>% dplyr::select(protein_names))
ggcorrplot(mcor, hc.order = TRUE, type = "lower", 
           outline.col = "lightgray",
           colors = c("#6D9EC1", "white", "#E46726"))
```

# Regression Analysis on Summarized Data

Classical differential analysis approach comparing median marker expressions (see [f1000 workflow](https://f1000research.com/articles/6-748/v2) by Nowicka et al.). 

## Plot Median Marker Expression

Plot all celltypes.

```{r median_expression_plot}
df_median = df_samples %>%
      group_by(file_name, donor, term, celltype) %>%
      summarise_at(protein_names, median)
df_median_long = gather(df_median, protein_name, median_expr, 
                        -file_name, -donor, -term, -celltype)
ggplot(df_median_long, aes(protein_name, median_expr, color = term)) + 
  geom_violin() + 
  facet_wrap(~celltype) +
  theme(axis.text.x  = element_text(angle = 90, vjust=0))
```

Zoom in on NK cells.

```{r median_expression_plot_nk}
df_median_long %<>% dplyr::filter(celltype == "NK")
ggplot(df_median_long, aes(term, median_expr, color = term)) + 
  geom_violin() + 
  geom_jitter(width = 0.2, alpha = 0.5) + 
  facet_wrap(~ protein_name, nrow = 2) +
  theme(axis.text.x = element_blank()) +
  ggtitle("NK")
```

Zoom in on marker pSTAT1.

```{r median_expression_plot_pstat1}
ggplot(df_median, aes(term, pSTAT1, color = term)) + 
  geom_violin() + 
  geom_jitter(width = 0.2, alpha = 0.5) + 
  facet_wrap(~celltype, nrow = 2) +
  theme(axis.text.x = element_blank())  
```

## Linear Mixed Model

Mixed model with median expression as response variable, experimental condition as explanatory variable, and donor as random effect. Fit separate models for each protein and celltype combination.

```{r median_expression_fit_separate}
calc_pvalue = function(fit) {
  summ = summary(fit)
  coefficients(summ)["term3rd trimester", "Pr(>|t|)"]
}
df_median_long = gather(df_median, protein_name, median_expr, 
                        -file_name, -donor, -term, -celltype)
df_fits = df_median_long %>% 
  group_by(protein_name, celltype) %>%
  nest() %>% 
  mutate(fit = map(data, ~ lmer(median_expr ~ term + (1|donor), .))) %>%
  mutate(pvalue_unadj = map_dbl(fit, ~ calc_pvalue(.))) %>%
  mutate(pvalue_adj = p.adjust(pvalue_unadj, method = "BH")) %>%
  dplyr::select(protein_name, celltype, pvalue_adj)
df_fits %>% 
  dplyr::filter(pvalue_adj < 0.05) %>% 
  dplyr::arrange(celltype) %>%
  print(n = Inf)
```

# Regression Analysis on All The Data

For the regression analysis, we will focus only on NK cells. We can repeat the same analysis for each cell type of interest.

## Generalized Linear Mixed Model

Fit a Generalized Linear Mixed Model (GLMM) with donor random effects. This function is a wrapper around the package `mbest` [@perry2017fast].

```{r glmm_fit}
df_samples_subset$term %<>% factor(levels = c("1st trimester",
                                              "3rd trimester"))
glmm_fit = CytoGLMM::cytoglmm(df_samples_subset, 
                              protein_names = protein_names,
                              condition = "term", group = "donor")
glmm_fit
plot(glmm_fit)
summary(glmm_fit) %>% dplyr::filter(pvalues_adj < 0.05)
```

## Generalized Linear Model with Bootstrap

Instead of modeling the donor effect, we can use bootstrap resampling. In our experience, this type of regression gives also good results when samples are not matched between conditions on the same donor.

```{r glm_fit}
glm_fit = CytoGLMM::cytoglm(df_samples_subset, 
                            num_boot = 1000,
                            protein_names = protein_names,
                            condition = "term", group = "donor")
glm_fit
plot(glm_fit)
summary(glm_fit) %>% dplyr::filter(pvalues_adj < 0.05)
```

## Mixture of Regressions

Fit a mixture of regression model to identity clusters of donors or outliers. This function is a wrapper around the package `flexmix` [@grun2007fitting].

```{r mix_fit}
num_donors = nlevels(as.factor(df_samples_subset$donor))
mix_fit = CytoGLMM::cytoflexmix(df_samples_subset, 
                                protein_names = protein_names,
                                condition = "term", group = "donor", 
                                ks = 1:num_donors)
plot(mix_fit)
```

The plotting function automatically uses the BIC criterion to select the number of clusters. In this case, it picks 10 clusters.

```{r bic_model_selection, fig.small=TRUE}
plot_model_selection(mix_fit)
```

# SummarizedExperiment

We create a ``SummarizedExperiment`` object containing marker, sample table, and untransformed protein counts. This way we can store all the information of this experiment in one file and load it again in subsequent analyses.

```{r create_summarized_experiment}
markers %<>% dplyr::filter(type != "none")
d_combined = df_samples %>% 
  select(markers$protein_name) %>%
  dplyr::mutate_all(.funs = inv_func) %>%
  dplyr::mutate_all(.funs = round) %>%
  as.matrix
row_data = df_samples %>% 
  select(sample_info_names) %>% 
  as.data.frame
col_data = markers %>% as.data.frame
se_aghaeepour2017immune = SummarizedExperiment(
  assays = list(exprs = d_combined),
  colData = col_data,
  rowData = row_data
)
save(se_aghaeepour2017immune, file = "se_aghaeepour2017immune.Rdata")
```

# Session Info {.unnumbered}

```{r session_info}
sessionInfo()
```

# References {.unnumbered}
